Overlapping Batches for the Assessment of Solution Quality in Stochastic Programs

### 2013-10-24 omrp.tex
streamlining the intro - I

We wish to assess the quality of a candidate solution $\xh \in X$. 
We define its quality by its optimality gap $\e{f(\xh,\xit)} - \zs$. 
The lower the optimality gap, the higher the quality of the solution and a zero optimality gap implies that the solution is optimal. 
Note that the candidate solution $\xh \in X$ may have been found in any way. 
We assume this solution is given.

Assessing solution quality is important in practice since (\ref{eq:sto_prog}) typically cannot be solved exactly, so one only has an approximate solution $\xh$ without verification of its quality.
Assessing solution quality is also a critical component of stopping criteria in algorithms.

In stochastic programming, an approximation of \ref{eq:sto_prog} is often generated by sampling from the distribution of $\xi$.
Let $\xit^1, \xit^2, \dots, \xit^m$ be an independent and identically distributed (iid) sample from the distribution of $\xit$.
A sampling approximation of (\ref{eq:sto_prog}) is given by
\begin{align} \tag{SP$_m$} \label{eq:sto_prog_m}
	z_m^* & = \min_{\x \in X} \frac{1}{m} \sum_{i=1}^m f(\x,\xiti).
\end{align}
We denote an optimal solution to (\ref{eq:sto_prog}) as $\xs$ and an optimal solution to (\ref{eq:sto_prog_m}) as $\xs_m$.

In many fields of optimization, it is common to evaluate an upper bound on the optimality gap through relaxations such as integrality or Lagrangian relations.

This motivates estimating the upper bound on the optimality gap by using Monte Carlo sampling.
Given a sample size $m$, an upper bound on the optimality gap can be obtained by $\e{f(\xh,\xit)} - \zs \leq \e{f(\xh,\xit)} - \e{\zs_m}$ due to the inequality $\e{\zs_m} \leq \zs$.
This bound improves as the sample size increases, that is, $\e{\zs_m} \leq \e{\zs_{m+1}} \leq \zs$ \citep{Mak1999,norkin_pflug_ruszczynski_98}.  
A straightforward estimate of $\e{f(\xh,\xit)}$ is the sample mean, $\frac{1}{m} \sum_{i=1}^m f(\xh,\xiti)$. 
Instead of $\e{\zs_m}$, simply $\zs_m$ can be used.  
The resulting point estimator is  $G_m(\xh) = \frac{1}{m} \sum_{i=1}^m f(\xh,\xiti) - \zs_m$.  
Here, we assume the same observations $\xit^1, \dots, \xit^m$ are used in both terms of $G_m(\xh)$.

Computing $G_m(\xh)$ involves solving an optimization problem (\ref{eq:sto_prog_m}) to obtain a lower bound estimator $\zs_m$, which complicates the statistical analysis. 
To enable statistical inference, the \emph{Multiple Replications Procedure} (MRP) of Mak et.\ al.\ \citep{Mak1999} generates $k$ independent estimators of $G_m(\xh)$, each using sample size $m$ and averages them to obtain a point estimator.  
To form confidence intervals, the sample variance of these estimators is calculated.  
This is essentially a nonoverlapping batch means type estimator ($k$ batches of size $m$) commonly used in simulation \citep{law_07}. 
We further review MRP estimators in \S \ref{ssec:mrp}.

>>>>

Unless $f$ has a simple structure or the number of realizations of $\xi$ is small, (\ref{eq:sto_prog}) typically cannot be solved exactly. 
An approximation of (\ref{eq:sto_prog}) is often generated by sampling from the distribution of $\xi$.
While other sampling schemes are possible, let $\xit^1, \xit^2, \dots, \xit^m$ be an independent and identically distributed (iid) sample from the distribution of $\xit$.
A sampling approximation of (\ref{eq:sto_prog}), often called the Sample Average Approximation (SAA), is given by
\begin{align} \tag{SP$_m$} \label{eq:sto_prog_m}
	z_m^* & = \min_{\x \in X} \frac{1}{m} \sum_{i=1}^m f(\x,\xiti).
\end{align}
We denote an optimal solution to (\ref{eq:sto_prog}) as $\xs$ and an optimal solution to (\ref{eq:sto_prog_m}) as $\xs_m$.

We wish to assess the quality of a candidate solution $\xh \in X$. 
Assessing solution quality is important in practice because as (\ref{eq:sto_prog}) typically cannot be solved exactly, one only has an approximate solution $\xh$ without verification of its quality.
Assessing solution quality is also a critical component of stopping criteria in algorithms designed to (approximately) solve (\ref{eq:sto_prog}).
We define the quality of $\xh$ by its optimality gap, $\e{f(\xh,\xit)} - \zs$. 
The lower the optimality gap, the higher the quality of the solution and a zero optimality gap implies that the solution is optimal. 
Note that the candidate solution $\xh \in X$ may have been found in any way. 
We assume this solution is given throughout this paper.

The optimality gap $\e{f(\xh,\xit)} - \zs$ cannot be calculated exactly. First,  for a given candidate solution $\xh \in X$, evaluation of $\e{f(\xh,\xit)}$ often involves a difficult multidimensional integral. 
Second, we do not know $\zs$. 
In many fields of optimization, the second problem is alleviated by obtaining lower bounds on $\zs$ through relaxations such as integrality, Lagrangian or semidefinite relations.
As a result, it is common to evaluate an upper bound on the optimality gap. 
This motivates estimating the upper bound on the optimality gap in stochastic programming by using Monte Carlo simulation.
Given a sample size $m$, an upper bound on the optimality gap can be obtained by $$
\e{f(\xh,\xit)} - \zs \leq \e{f(\xh,\xit)} - \e{\zs_m}
$$ 
due to the inequality $\e{\zs_m} \leq \zs$.
This bound improves as the sample size increases, that is, $\e{\zs_m} \leq \e{\zs_{m+1}} \leq \zs$ \citep{Mak1999,norkin_pflug_ruszczynski_98}.  
A straightforward estimate of $\e{f(\xh,\xit)}$ is the sample mean, $\frac{1}{m} \sum_{i=1}^m f(\xh,\xiti)$. 
Instead of $\e{\zs_m}$, simply $\zs_m$ can be used.  
The resulting point estimator of optimality gap is then 
$$
G_m(\xh) = \frac{1}{m} \sum_{i=1}^m f(\xh,\xiti) - \zs_m.
$$ 
Here, we assume the same observations $\xit^1, \dots, \xit^m$ are used in both terms of $G_m(\xh)$.

Computing $G_m(\xh)$ involves solving an optimization problem (\ref{eq:sto_prog_m}) to obtain a lower bound estimator of $\zs$, $\zs_m$, which complicates the statistical analysis. 
As mentioned above, to enable statistical inference, the MRP of \citet{Mak1999} generates $k$ independent estimators of $G_m(\xh)$, each using sample size $m$ ($k$ batches of size $m$) and averages them to obtain a point estimator.  
To form confidence intervals, the sample variance of these estimators is calculated.  
%This is essentially a nonoverlapping batch means type estimator ($k$ batches of size $m$) commonly used in simulation \citep{law_07}. 
(We further review MRP estimators in \S \ref{ssec:mrp}.)
